Bayesian multivariate skew-normal finite mixture model for analysis of infant development trajectories

### 2019-06-07 manuscript/manuscript.tex
algo

Mixtures of multivariate non-symmetric distributions such as the multivariate skew-normal (MSN) distribution allow for the nuances of the marginal density to be captured with a more parsimonious set of mixture components. Mixtures of MSN distributions have been dealt with previously in a Bayesian context (Fr\"{u}hwirth-Schnatter \& Pyne, 2010), however in these models, focus lies primary on marginal density estimation and inference on the mixture components (i.e. clusters) is not discussed. More recently, the mixtures of skew-$t$ factor analysis (MSTFA) model has been proposed for settings in which cluster-specific inference is of primary interest (Lin \textit{et al.} 2018). However, an important feature not included in the MSTFA is the ability to explain individual-level cluster membership as a function of covariates of interest. Additionally, parameter estimation proposed by Lin et al. for the MSTFA relies on a prohibitively complex EM algorithm and does not enjoy the inferential benefits of a Bayesian approach, namely the ability to incorporate prior information into a model and make posterier probability statements. Our proposed model improves on these previous works by estimating parameters in a Bayesian framework as well as including the ability to fit a multinomial logit regression to cluster membership probabilities using a novel application of data augmentation with the P\'olya Gamma distribution.

@@

Let $\mathbf{y}_{i}$ be the $J \times 1$ observation vector for subject $i$ such that $y_{ij}$ is the observation for subject $i$ at timepoint $j$. We assume for now that $\mathbf{y}_{i}$ is fully observed. Later, we relax this assumptin by allowing for missingness in the components of $\mathbf{y}_{i}$. The general formulation of the multivariate skew normal mixure model is as follows.

@@

A primary concern of our model is with identification of latent infant development clusters. We accomplish this via multinomial logit regression model on cluster membership, which utilizes P\'{o}lya-Gamma data-aumentation to allow for updating of all parameters using Gibbs sampling. The multinomial logit model is as follows for $k = 1,...,K$.

@@

\item Emphasize that PG data augmentation for the multinomial model results in a PG mixure of experts model, which is a computationally efficient way to model edge weights.

@@

\subsection{Simulation to Assess Sensitivity to Mispecified K}

@@

We now dervive the joint posterior distribution of the parameters $\mathbf{B}^*$ and $\boldsymbol\Sigma$.
\begin{align*} 
p(\mathbf{B}^*,\boldsymbol\Sigma|\mathbf{X}^*,\mathbf{Y}) & \propto  p(\mathbf{Y}|\mathbf{X}^*,\mathbf{B}^*,\boldsymbol\Sigma)p(\mathbf{B}^*|\boldsymbol\Sigma)p(\boldsymbol\Sigma)\\
&  \propto |\boldsymbol\Sigma|^{-n/2}\exp \left \{ -\frac{1}{2} \tr \left [(\mathbf{Y} - \mathbf{X^*}\mathbf{B}^*)^T(\mathbf{Y} - \mathbf{X^*}\mathbf{B}^*)\boldsymbol\Sigma^{-1} \right ]\right \} \\
& \times |\boldsymbol\Sigma|^{-(P+1)/2} \exp \left \{ -\frac{1}{2} \tr \left [(\mathbf{B}^* - \mathbf{B}^*_0)^T \mathbf{L}_0(\mathbf{B}^* - \mathbf{B}^*_0)\boldsymbol\Sigma^{-1} \right ]\right \} \\
& \times |\boldsymbol\Sigma|^{(\nu_0 + J + 1)/2} \exp \left \{ -\frac{1}{2} \tr (\mathbf{V}_0 \boldsymbol\Sigma^{-1}) \right \}
\end{align*}

@@

The multivariate normal conditional imputation derivations are given for a single cluster without loss of generality. In practive, the data and parameters in this section would be replaced by cluster specific estimates in the case of clustering.

@@

The blockwise inversion formula was used to invert $\boldsymbol\Sigma$ according to the following reparameterizations.
\begin{align*}
\boldsymbol\Sigma_{11}^* & = \boldsymbol\Sigma_{11}^{-1} + \boldsymbol\Sigma_{11}^{-1}\boldsymbol\Sigma_{12}(\boldsymbol\Sigma_{22} - \boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\boldsymbol\Sigma_{12})^{-1}\boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\\
\boldsymbol\Sigma_{12}^* & = -\boldsymbol\Sigma_{11} \boldsymbol\Sigma_{12}(\boldsymbol\Sigma_{22}-\boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\boldsymbol\Sigma_{12})^{-1}\\
\boldsymbol\Sigma_{21}^* & = -(\boldsymbol\Sigma_{22} - \boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\boldsymbol\Sigma_{12})^{-1}\boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\\
\boldsymbol\Sigma_{22}^* & = (\boldsymbol\Sigma_{22} - \boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\boldsymbol\Sigma_{12})^{-1}     
\end{align*}

@@

@@

@@

>>>>

Mixtures of multivariate non-symmetric distributions such as the multivariate skew-normal (MSN) distribution allow for the nuances of the marginal density to be captured with a more parsimonious set of mixture components. Mixtures of MSN distributions have been dealt with previously in a Bayesian context (Fr\"{u}hwirth-Schnatter \& Pyne, 2010), however in these models, focus lies primary on marginal density estimation and inference on the mixture components (i.e. clusters) is not discussed. More recently, the mixtures of skew-$t$ factor analysis (MSTFA) model has been proposed for settings in which cluster-specific inference is of primary interest (Lin \textit{et al.} 2018). However, an important feature not included in the MSTFA is the ability to explain individual-level cluster membership as a function of covariates of interest. Additionally, parameter estimation proposed by Lin et al. for the MSTFA relies on a prohibitively complex EM algorithm and does not enjoy the inferential benefits of a Bayesian approach, namely the ability to incorporate prior information into a model and make posterior probability statements. Our proposed model improves on these previous works by estimating parameters in a Bayesian framework as well as including the ability to fit a multinomial logit regression to cluster membership probabilities using a novel application of data augmentation with the P\'olya Gamma distribution.

@@

Let $\mathbf{y}_{i}$ be the $J \times 1$ observation vector for subject $i$ such that $y_{ij}$ is the observation for subject $i$ at timepoint $j$. We assume for now that $\mathbf{y}_{i}$ is fully observed. Later, we relax this assumption by allowing for missingness in the components of $\mathbf{y}_{i}$. The general formulation of the multivariate skew normal mixture model is as follows.

@@

A primary concern of our model is with identification of latent infant development clusters. We accomplish this via multinomial logit regression model on cluster membership, which utilizes P\'{o}lya-Gamma data-augmentation to allow for updating of all parameters using Gibbs sampling. The multinomial logit model is as follows for $k = 1,...,K$.

@@

\item Emphasize that PG data augmentation for the multinomial model results in a PG mixture of experts model, which is a computationally efficient way to model edge weights.

@@

\subsection{Simulation to Assess Sensitivity to Misspecified K}

@@

We now derive the joint posterior distribution of the parameters $\mathbf{B}^*$ and $\boldsymbol\Sigma$.
\begin{align*} 
p(\mathbf{B}^*,\boldsymbol\Sigma|\mathbf{X}^*,\mathbf{Y}) & \propto  p(\mathbf{Y}|\mathbf{X}^*,\mathbf{B}^*,\boldsymbol\Sigma)p(\mathbf{B}^*|\boldsymbol\Sigma)p(\boldsymbol\Sigma)\\
&  \propto |\boldsymbol\Sigma|^{-n/2}\exp \left \{ -\frac{1}{2} \tr \left [(\mathbf{Y} - \mathbf{X^*}\mathbf{B}^*)^T(\mathbf{Y} - \mathbf{X^*}\mathbf{B}^*)\boldsymbol\Sigma^{-1} \right ]\right \} \\
& \times |\boldsymbol\Sigma|^{-(P+1)/2} \exp \left \{ -\frac{1}{2} \tr \left [(\mathbf{B}^* - \mathbf{B}^*_0)^T \mathbf{L}_0(\mathbf{B}^* - \mathbf{B}^*_0)\boldsymbol\Sigma^{-1} \right ]\right \} \\
& \times |\boldsymbol\Sigma|^{(\nu_0 + J + 1)/2} \exp \left \{ -\frac{1}{2} \tr (\mathbf{V}_0 \boldsymbol\Sigma^{-1}) \right \}
\end{align*}

@@

The multivariate normal conditional imputation derivations are given for a single cluster without loss of generality. In practice, the data and parameters in this section would be replaced by cluster specific estimates in the case of clustering.

@@

The block-wise inversion formula was used to invert $\boldsymbol\Sigma$ according to the following reparameterizations.
\begin{align*}
\boldsymbol\Sigma_{11}^* & = \boldsymbol\Sigma_{11}^{-1} + \boldsymbol\Sigma_{11}^{-1}\boldsymbol\Sigma_{12}(\boldsymbol\Sigma_{22} - \boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\boldsymbol\Sigma_{12})^{-1}\boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\\
\boldsymbol\Sigma_{12}^* & = -\boldsymbol\Sigma_{11} \boldsymbol\Sigma_{12}(\boldsymbol\Sigma_{22}-\boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\boldsymbol\Sigma_{12})^{-1}\\
\boldsymbol\Sigma_{21}^* & = -(\boldsymbol\Sigma_{22} - \boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\boldsymbol\Sigma_{12})^{-1}\boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\\
\boldsymbol\Sigma_{22}^* & = (\boldsymbol\Sigma_{22} - \boldsymbol\Sigma_{21}\boldsymbol\Sigma_{11}^{-1}\boldsymbol\Sigma_{12})^{-1}     
\end{align*}

@@

\bibitem{ } Chen JT, Gupta AK. Matrix variate skew normal distributions. \textit{Statistics}. 2005 Jun 1;39(3):247-53.

@@

\bibitem{ } Viroli C. Finite mixtures of matrix normal distributions for classifying three-way data. \textit{Statistics and Computing}. 2011 Oct 1;21(4):511-22.

@@

\bibitem{ } Zeller CB, Cabral CR, Lachos VH. Robust mixture regression modeling based on scale mixtures of skew-normal distributions. \textit{Test}. 2016 Jun 1;25(2):375-96.
